# Economic and Health Impact of Severe Weather Events in the USA

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.  

Your data analysis must address the following questions:  

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
  
2. Across the United States, which types of events have the greatest economic
consequences?  
  
- The analysis document must have at least one figure containing a plot.  
  
- Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.  
   
   
## Data Processing

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size, from the following web site:  
  
- Storm Data: *https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2*   
  

There is also some documentation of the database available for how some of the variables are constructed/defined: 
  
- National Weather Service Storm Data Documentation: *https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf*   
  
- National Climatic Data Center Storm Events FAQ: *https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf* 
  
  
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


### Loading the data
  
  
The data for this assignment can be found at the following url as a bzip2 compressed file containing the `repdata-data-StormData.csv` file in csv file format:

- Url: *ttps://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2*
  
- File: *repdata-data-StormData.csv* 
  
    
The following code will be used to download and read the data into the `StormData`
data frame:

```{r cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp, method="curl")
StormData <- read.csv(bzfile(temp, "repdata-data-StormData.csv"), stringsAsFactors = FALSE)
unlink(temp)
rm(temp)
```
  
  
### Analyzing data structure

Let's take a look at the structure of the data loaded:

```{r}
str(StormData)
```

We notice that there are 902.297 observations with 37 variables. 
  
>   **Not all variables are of interest in this study!** 
  
  We shall use only the following variables that we consider relevant to this study:
  
  - **BGN_DATE** (column 2): Character value with the date in which the event took place in the **MM/DD/YEAR HH:MM:SS** format. We may want to have a look at the evolution of impact in population health and economy with time or at each year.
  
  - **EVTYPE** (column 8): Character value with the type of the event, since we want to explore how each type of event impacts population health and economy. The official table of events is described in item 2.1.1 on page 6 of the [document describing the data](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).
  
  - **FATALITIES** and **INJURIES** (columns 23 and 24): Numeric values with the number of fatalities and injuries caused by each event, once we will measure the impact of each type of event in population health by the sum of these two variables.
  
  - **PROPDMG** and **CROPDMG** (columns 25 and 27): Numeric values with an estimate of the value of damages, in properties and in crops, caused by each event, once we will consider measure of the economic impact of each type of event by the sum of these two variables.
  
  - **PROPDMGEXP** and **CROPDMGEXP** (columns 26 and 28): Character values with an alphabetical character suffix to PROPDMG and CROPDMG  signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions.
  
  - **REFNUM** (column 37): Numeric value with an index numner to the specific weather event, just to be able to track back to the event, if necessary.
  
  The remaining variables will not be used in this particular study.
  
  
### Subsetting the data  
  
  Before we continue digging deeper in the data, we will extract from the data loaded only the rows and columns of interest.
  
  We will create two new data frames:
  
  - **HealthImpact** with data relevant for population health: **BGN_DATE** (column 2), **EVTYPE** (column 8), **FATALITIES** and **INJURIES** (columns 23 and 24)
  
  - **EconomicImpact** with data relevant to economic impact: **BGN_DATE** (column 2), **EVTYPE** (column 8), **PROPDMG** (column 25), **PROPDMGEXP** (columns 26), **CROPDMG** (columns 27), and **CROPDMGEXP** (columns 28)
  
  For that we will use the following code:
  
```{r cache=TRUE}
HealthImpact<-StormData[StormData$INJURIES > 0 | StormData$FATALITIES > 0, c(2, 8, 23, 24, 37)]

EconomicImpact<-StormData[StormData$PROPDMG > 0 | StormData$CROPDMG > 0, c(2, 8, 25, 26, 27, 28, 37)]

# Remove the big StormData file
# rm(StormData)
```
  
  Let's also adjust some of the columns:
  
  - The character columns for EVTYPE, PROPDMGEXP e CROPDMGEXP to factor variables
  
  - The BGN_DATE to date format 
  
```{r}
HealthImpact$EVTYPE<-as.factor(HealthImpact$EVTYPE)

EconomicImpact$EVTYPE<-as.factor(EconomicImpact$EVTYPE)

EconomicImpact$PROPDMGEXP<-as.factor(EconomicImpact$PROPDMGEXP)

EconomicImpact$CROPDMGEXP<-as.factor(EconomicImpact$CROPDMGEXP)

HealthImpact$BGN_DATE<-as.POSIXct(strptime(HealthImpact$BGN_DATE, format = "%m/%d/%Y %H:%M:%S", tz ="GMT"))

EconomicImpact$BGN_DATE<-as.POSIXct(strptime(EconomicImpact$BGN_DATE, format = "%m/%d/%Y %H:%M:%S", tz ="GMT"))
```
  
### Health Impact Data
  
  Now let's analyze the structure and summary of the health impact subset data that will be used:
  
```{r}
str(HealthImpact)

summary(HealthImpact)
```
  
  For the Health Impact data:
  
  - We now have 21.929 observations, or severe weather events, with any fatalities or injuries among the 902.297 observations.
  
  - These severe weather events were observed between 1950-01-03 and 2011-11-30.
  
  - We do not have any missing values.  
  
  - The EVTYPE (event type) column has 220 levels, when it should have only 48 according to the table of events in the documentation.
  
  
### Event Types in Health Impact Data

  Let's explore the types of events types in our Health Impact data.
  
```{r}
as.data.frame(table(HealthImpact$EVTYPE))
``` 
  
  We can see that we have 220 different types of events, instead of 48, because of: 
  
  - Mistypings
  
  - Use of upper and lower cases
  
  - Failure to use the standard
  
  
  We will attempt to impprove the data by:
  - Converting type to upper case
  
  - Convert to HURRICANE any event type which contains TYPHOON or HURRICANE
  
  - Convert to TORNADO any event type which contains TORNADO
  
  - Convert to LIGHTNING any event type which contains LIGHTN
  
  - Convert to THUNDERSTORM any event type which contains THUNDER or TSTM
  
  - Convert to FLOOD any event type which contains FLOOD
  
  - Convert to HEAT any event type which contains HEAT
  
  - Convert to WILDFIRE any event type which contains FIRE
  
  - Convert to ICE STORM any event type which contains ICE
  
  - Convert to COLD any event type which contains COLD
  
  
```{r}
HealthImpact$EVTYPE<-toupper(HealthImpact$EVTYPE)
HealthImpact[grep("hurricane", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HURRICANE"
HealthImpact[grep("typho", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HURRICANE"
HealthImpact[grep("tornado", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"TORNADO"
HealthImpact[grep("lightn", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"LIGHTNING"
HealthImpact[grep("thunder", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"THUNDERSTORM"
HealthImpact[grep("tstm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"THUNDERSTORM"
HealthImpact[grep("flood", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FLOOD"
HealthImpact[grep("fld", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FLOOD"
HealthImpact[grep("heat", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAT"
HealthImpact[grep("fire", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"WILDFIRE"
HealthImpact[grep("ice", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"ICE STORM"
HealthImpact[grep("icy", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"ICE STORM"
HealthImpact[grep("cold", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("chill", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("therm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("low temp", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("rip cur", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"RIP CURRENT"
HealthImpact[grep("snow", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAVY SNOW"
HealthImpact[grep("warm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAT"
HealthImpact[grep("slide", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"DEBRIS FLOW"
HealthImpact[grep("rain", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAVY RAIN"
HealthImpact[grep("freez", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FROST FREEZE"
HealthImpact[grep("frost", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FROST FREEZE"
HealthImpact[grep("wind", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"STRONG/HIGH WIND"
HealthImpact[grep("wint", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"WINTER STORM/WEATHER"
HealthImpact[grep("hail", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HAIL"
HealthImpact[grep("avalanc", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"AVALANCHE"
HealthImpact[grep("tropical storm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"TROPICAL STORM"
HealthImpact[grep("surge", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"STORM SURGE/TIDE"
HealthImpact[grep("surf", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HIGH SURF"

as.data.frame(table(HealthImpact$EVTYPE))
``` 
  
  Now we have 49 different event types, much closer to the 48 we would expect, even though some types of events do not appear on the official list.
  
  
### Total Casualties

  In order to evaluate the impact of different types of severe weather on population health, we will compute the total number of casualties as the sum of the number of fatalities and injuries.
  
```{r}
HealthImpact$CASUALTIES<-HealthImpact$INJURIES+HealthImpact$FATALITIES
```

### Economic Impact Data
  
  Now let's analyze the structure and summary of the economic impact subset data that will be used:
  
```{r}
str(EconomicImpact)

summary(EconomicImpact)
```
  
  For the Health Impact data:
  
  - We now have 245.031 observations, or severe weather events, with an estimated value of damages in properties or crops.
  
  - These severe weather events were observed between 1950-01-03 and 2011-11-30.
  
  - We do not have any missing values.  
  
  - The EVTYPE (event type) column has 431 levels, when it should have only 48 according to the table of events in the documentation.
  
  - The PROPDMGEXP and CROPDMGEXP columns have 16 and 8 levels respectively while we expected a much lower number since these suffixes should stand for Billions, Millions or Thousands of dollars.
  
  
### Event Types in Economic Impact Data

  Let's explore the types of events types in our Health Impact data.
  
```{r}
as.data.frame(table(EconomicImpact$EVTYPE))
``` 
  
  We can see that we have 431 different types of events, instead of 48, because of: 
  
  - Mistypings
  
  - Use of upper and lower cases
  
  - Failure to use the standard
  
  
  We will attempt to impprove the data by applying the same conversions we applied to the Health Impact Data:
  
  - Converting type to upper case
  
  - Convert to HURRICANE any event type which contains TYPHOON or HURRICANE
  
  - Convert to TORNADO any event type which contains TORNADO
  
  - Convert to LIGHTNING any event type which contains LIGHTN
  
  - Convert to THUNDERSTORM any event type which contains THUNDER or TSTM
  
  - Convert to FLOOD any event type which contains FLOOD
  
  - Convert to HEAT any event type which contains HEAT
  
  - Convert to WILDFIRE any event type which contains FIRE
  
  - Convert to ICE STORM any event type which contains ICE
  
  - Convert to COLD any event type which contains COLD
  
  
```{r}
HealthImpact$EVTYPE<-toupper(HealthImpact$EVTYPE)
HealthImpact[grep("hurricane", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HURRICANE"
HealthImpact[grep("typho", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HURRICANE"
HealthImpact[grep("tornado", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"TORNADO"
HealthImpact[grep("lightn", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"LIGHTNING"
HealthImpact[grep("thunder", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"THUNDERSTORM"
HealthImpact[grep("tstm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"THUNDERSTORM"
HealthImpact[grep("flood", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FLOOD"
HealthImpact[grep("fld", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FLOOD"
HealthImpact[grep("heat", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAT"
HealthImpact[grep("fire", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"WILDFIRE"
HealthImpact[grep("ice", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"ICE STORM"
HealthImpact[grep("icy", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"ICE STORM"
HealthImpact[grep("cold", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("chill", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("therm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("low temp", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"COLD"
HealthImpact[grep("rip cur", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"RIP CURRENT"
HealthImpact[grep("snow", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAVY SNOW"
HealthImpact[grep("warm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAT"
HealthImpact[grep("slide", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"DEBRIS FLOW"
HealthImpact[grep("rain", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HEAVY RAIN"
HealthImpact[grep("freez", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FROST FREEZE"
HealthImpact[grep("frost", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"FROST FREEZE"
HealthImpact[grep("wind", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"STRONG/HIGH WIND"
HealthImpact[grep("wint", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"WINTER STORM/WEATHER"
HealthImpact[grep("hail", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HAIL"
HealthImpact[grep("avalanc", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"AVALANCHE"
HealthImpact[grep("tropical storm", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"TROPICAL STORM"
HealthImpact[grep("surge", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"STORM SURGE/TIDE"
HealthImpact[grep("surf", HealthImpact$EVTYPE, ignore.case = TRUE, value=FALSE),2]<-"HIGH SURF"

as.data.frame(table(HealthImpact$EVTYPE))
``` 
  
  Now we have 49 different event types, much closer to the 48 we would expect, even though some types of events do not appear on the official list.

### Adjusting the Magnitude Suffixes

  As we saw, the PROPDMGEXP and CROPDMGEXP (magnitude of PROPDMG and CROPDMG) columns have 16 and 8 levels respectively, which are numbers larger than we would expect. 
  
  We expected to have the following suffixes:
  
  - **B** for billions of dollars
  
  - **M** for millions of dollars
  
  - **K** for thousands of dollars
  
  We need than to take a further look into those suffixes.
  
  Let's create a table with the different suffixes found for PROPDMGEXP and CROPDMGEXP, when the values of PROPDMG and CROPDMG that are greater than zero: 
  
```{r}
# Data for PROPDMG
x<-table(StormData$PROPDMGEXP, StormData$PROPDMG > 0)
dimnames(x)[[2]]<-c("PROPDMG=0","PROPDMG>0")
x

# Data for CROPDMG
y<-table(StormData$CROPDMGEXP, StormData$CROPDMG > 0)
dimnames(y)[[2]]<-c("CROPDMG=0","CROPDMG>0")
y
```


### Adjusting Types of Events

  The number of event types present in the data (488) is much larger than what is described in the documentation, so we need to take a further look into it to 
  
  
### Preprocessing the data

In order to be able to work with date and time functions to aggregate and plot the data, a new variable, calles `datetime` will be created in the `activity` data frame with the date & time of each of the 5 minutes intervals, in the POSIXct format:

```{r}
#activity$datetime<-as.POSIXct(strptime(paste(activity$date, formatC(activity$interval, width=4, flag="0")), "%Y-%m-%d %H%M"))
```


## Results

### Impact in Population Health

For this part of the assignment we will:
  

### Impact in the Economy

For this part of the assigment, we will:

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

1. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

